---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "75%",
  fig.align = "center"
)
```

# {Lcpp}: Convenient forward algorithm in C++

This package contains convenient R-wrapper functions of the forward algorithm used to estimate hidden Markov models (HMMs) via direct numerical maximum likelihood estimation. The algorithm calculates the negative log-likelihood recursively as a matrix product (see [Zucchini et al. 2016](https://www.taylorfrancis.com/books/mono/10.1201/b20790/hidden-markov-models-time-series-walter-zucchini-iain-macdonald-roland-langrock)). Thus, implementation in C++ offers 10-20 times faster evaluation times. Several versions are contained, including homogeneous HMMs and a general implementation with a pre-calculated array of transition probability matrices (t.p.m.s). These functions can then be included in the negative log-likelihood function, once the initial distribution (delta), the allprobs matrix and the t.p.m.(s) are calculated.

Various different implementations will be added as needed. Have fun!

## Installation

```{r, eval = FALSE}
devtools::install_github("janoleko/Lcpp")
```