---
title: "Periodic HMMs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Periodic HMMs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ref.bib
link-citations: yes
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "img/",
  fig.align = "center",
  fig.dim = c(8, 6),
  out.width = "85%"
)
library("Lcpp")
```

This example shows how to fit an HMM with the state process being a periodically inhomogeneous Markov chain, which means formally that for all $t$

$$
\Gamma^{(t+L)} = \Gamma^{(t)},
$$
where $\Gamma^{(t)}$ is the transition probability matrix at time $t$ and $L$ is the cycle length. This can conveniently modeled by letting the off-diagonal elements be trigonometric functions of a cyclic variable such as time of day.

#### Setting parameters for simulation
```{r parameters}
# parameters
mu = c(4, 14)
sigma = c(3, 5)

tod = rep(1:24, 100)
L = 24
beta = matrix(c(-1, 3, -4, 
                -2, -2, 1), nrow = 2, byrow = TRUE)
Gamma = tpm_p(seq(1, 24, by = 1), L, beta)
Delta = stationary_p(Gamma)

# having a look at the periodically stationary distribution
color = c("orange", "deepskyblue")
plot(Delta[,1], type = "l", lwd = 3, col = color[1], bty = "n")
points(Delta[,1], pch = 19, col = color[1])
lines(Delta[,2], lwd = 3, col = color[2])
points(Delta[,2], pch = 19, col = color[2])
```

# Simulating data
```{r data}
# simulation
z = rep(1:24, 200) # time of day variable
n = length(z)
set.seed(123)
s = x = rep(NA, n)
s[1] = sample(1:2, 1, prob = Delta[z[1],])
x[1] = stats::rnorm(1, mu[s[1]], sigma[s[1]])
for(t in 2:n){
  s[t] = sample(1:2, 1, prob = Gamma[s[t-1],,z[t]])
  x[t] = stats::rnorm(1, mu[s[t]], sigma[s[t]])
}

par(mfrow = c(2,1))
plot(x[1:400], bty = "n", pch = 20, ylab = "x", 
     col = c(color[1], color[2])[s[1:400]])
boxplot(x ~ z, xlab = "time of day")
# we see a periodic pattern in the data
```

#### Writing the negative log-likelihood function

```{r mllk}
mllk = function(theta.star, x, z){
  beta = matrix(theta.star[1:6], nrow = 2)
  Gamma = tpm_p(beta = beta) # calculating all L tpms
  delta = stationary_p(Gamma, t = z[1]) # periodically stationary HMM
  mu = theta.star[7:8]
  sigma = exp(theta.star[9:10])
  # calculate all state-dependent probabilities
  allprobs = matrix(1, length(x), 2)
  for(j in 1:2){ allprobs[,j] = stats::dnorm(x, mu[j], sigma[j]) }
  # return negative for minimization
  -forward_p(delta, Gamma, allprobs, z)
}
```

#### Fitting an HMM to the data
```{r model, warning=FALSE}
theta.star = c(-1,-1, rep(0, 4), # starting values state process
               3, 10 ,log(2),log(6)) # starting values state-dependent process
s = Sys.time()
mod = stats::nlm(mllk, theta.star, x = x, z = z)
Sys.time()-s
```
Really fast!

#### Visualizing results
Again, we use `tpm()` and `stationary()` to tranform the unconstraint parameters to working parameters.
```{r visualization}
# transform parameters to working
beta_hat = matrix(mod$estimate[1:6], nrow = 2)
Gamma_hat = tpm_p(beta = beta_hat)
Delta_hat = stationary_p(Gamma_hat)
mu_hat = mod$estimate[7:8]
sigma_hat = exp(mod$estimate[9:10])

delta_hat = apply(Delta_hat, 2, mean)

par(mfrow = c(1,2))
hist(x, prob = TRUE, bor = "white", breaks = 40, main = "")
curve(delta_hat[1]*dnorm(x, mu_hat[1], sigma_hat[1]), add = TRUE, lwd = 2, 
      col = color[1], n=500)
curve(delta_hat[2]*dnorm(x, mu_hat[2], sigma_hat[2]), add = TRUE, lwd = 2, 
      col = color[2], n=500)
curve(delta_hat[1]*dnorm(x, mu_hat[1], sigma_hat[1])+
        delta_hat[2]*dnorm(x, mu[2], sigma_hat[2]),
      add = TRUE, lwd = 2, lty = "dashed", n = 500)
legend("topright", col = c(color[1], color[2], "black"), lwd = 2, bty = "n",
       lty = c(1,1,2), legend = c("state 1", "state 2", "marginal"))

plot(Delta_hat[,1], type = "l", lwd = 3, col = color[1], bty = "n", 
     xlab = "time of day", ylab = "probs")
points(Delta_hat[,1], pch = 19, col = color[1])
lines(Delta_hat[,2], lwd = 3, col = color[2])
points(Delta_hat[,2], pch = 19, col = color[2])

```