---
title: "State space models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{State space models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ref.bib
link-citations: yes
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "img/",
  fig.align = "center",
  fig.dim = c(8, 6),
  out.width = "85%"
)
library("Lcpp")
```

This vignette shows how to fit state space models which can be interpreted as generalization of HMMs to continuous state spaces. Several approaches exist to fitting such models, but @langrock2011some showed that very general state space models can be fitted via approximate maximum likelihood estimation, when the continous state space is finely discretized. Here, we will showcase this approach for a basic stochastic volatily model, which can be used to describe fincancial markets. In this model the unobserved marked volatility is described by an AR(1) process:

$$
g_t = \phi g_{t-1} + \sigma \eta_t, \qquad \eta_t \sim N(0,1),
$$
with autoregression parameter $\phi < 1$. We could then model share returns $y_t$ as
$$
y_t = \beta \epsilon_t \exp(g_t / 2),
$$
where $\beta > 0$ is the baseline standard deviation of the returns (when $g_t$ is in equilibrium). This implies
$$
y_t \mid g_t \sim N(0, (\beta e^{g_t / 2})^2).
$$

#### Simulating data from the stochastic volatility model

```{r data}
beta = 5 # baseline standard deviation
phi = 0.95 # AR parameter
sigma = 0.5 # variability of the AR process

n = 500
set.seed(123)
g = y = rep(NA, n)
g[1] = rnorm(1, 0, sigma / sqrt(1-phi^2)) # stationary distribution of AR process
y[1] = stats::rnorm(1, 0, beta*exp(g[1]/2)) 
# conditional distribution of y_1 given underlying volatility
for(t in 2:n){
  g[t] = rnorm(1, phi*g[t-1] , sigma) # transition density
  y[t] = stats::rnorm(1, 0, beta*exp(g[t]/2)) 
  # conditional distribution of y_t given underlying volatility
}

# share returns
plot(y, type = "l", bty = "n", ylim = c(-100,40))
# true underlying volatility
abline(h = -70)
lines(g*4-70, col = "orange", lwd = 2)
```

#### Writing the negative log-likelihood function

```{r mllk}
mllk = function(theta.star, y, bm, m){
  phi = plogis(theta.star[1])
  sigma = exp(theta.star[2])
  beta = exp(theta.star[3])
  b = seq(-bm, bm, length = m+1)
  h = b[2]-b[1]
  bstar = (b[-1] + b[-(m+1)])/2
  Gamma = sapply(bstar, dnorm, mean = phi*bstar, sd = sigma) * h
  Gamma = Gamma / rowSums(Gamma) # normalizing out approximation errors
  delta = h * dnorm(bstar, 0, sigma/sqrt(1-phi^2)) # stationary distribution
  # calculating state dependent density
  allprobs = t(sapply(y, dnorm, mean = 0, sd = beta * exp(bstar)))
  # return negative for minimization
  -forward(delta, Gamma, allprobs)
}
```

#### Fitting an SSM to the data

```{r model, warning=FALSE}
theta.star = c(qlogis(0.8), log(0.5), log(4))
bm = 5 # relevant range of underlying volatility (-5,5)
m = 100 # number of approximating states

t1 = Sys.time()
mod = stats::nlm(mllk, theta.star, y = y, bm = bm, m = m)
Sys.time()-t1
```

#### Results
```{r results}
(phi = plogis(mod$estimate[1]))
(sigma = exp(mod$estimate[2]))
(beta = exp(mod$estimate[3]))
```


#### References